Can historical data predict population responses to climate change experiments?

### 2016-12-20 precip/manuscript/precip_ms.tex
adding manuscript edits to results climate model section

The continuously recorded soil moisture data also recorded treatment effects on soil moisture, but these effects were weaker on average than the more complete spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}). We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring. The irrigated plots were 30\% wetter during the fall and summer but only 20 to 25\% wetter during the spring.  Treatment differences were slightly larger during rainy periods, especially in the spring. We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature within the plots.

The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest. Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature. During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content observed.

@@

The cover of \textit{H. comata} and \textit{P. spicata} fell significantly in the drought plots from the pre-treatment cover 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}). The cover of \textit{P. secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).  In contrast to the grasses, the cover of \textit{A. tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).

Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species. After comparing WAIC scores for the growth models with and without the size by treatment effects, we retained size by treatment interaction effects in the growth models for \textit{A. tripartita} and \textit{P. secunda}, and the survival model for \textit{P. secunda}. For \textit{A. tripartita} the model showed a significant size by treatment effect in the drought plots: the effect of drought was positive for plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls. \textit{A. tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level), irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size. \textit{H. comata} growth showed a strong (but not significant) decrease in the drought plots and no response to the irrigation plots.  Like \textit{A. tripartita}, \textit{P.secunda} growth also showed a size by treatment effect in the drought plots, with the negative effects on growth becoming greater for larger plants. \textit{P. secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although these effects were technically not significant). \textit{P. spicata} growth was relatively unaffected by the drought and irrigation treatments.

The effects of drought were negative and signficant on the survival of all three grass species (fig \ref{fig:)

We observed a number of strong correlations between the year effects in the vital rate models and the seasonal soil moisture variables and table x reports the strongest three correlations for each species and vital rate (table). In general the soil moisture during the spring and summer period appeared to be more strongly correlated with the year effects for each species than the fall soil moisture, but see \textit{P. spicata} vr.  All three time lags show up in the choosen variables, but 1 out of X variables choosen involved soil moisture in the year immediately before the first year of the transition, (subscript "0"), while variables involving soil moisture in the year before that (subscript "lag") and after that (subscript "1") were choosen less often.

We observed vital rate and species-specific effects of soil moisture on each of these species after fitting the vital rate models with the soil moisture covariates discussed above (fig). DISSCUSS significant soil moisture effects: for \textit{A. tripartita} growth was significantly affected by ...  For \textit{H. comata}.  For \textit{P. secunda}. And for \textit{P. spicata}. Soil moisture by size effects on \textit{A. tripartita} and \textit{P. secunda} showed that ()

@@

\section{Tables}

@@

\section{Figures}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{start_to_finish_cover_change}
	\caption{Log change in cover in each of the experimental plots from the pre-treatment monitoring in 2011 to the last year of the experiment in 2016. Box plots show the median cover change and the interquartile range.}
	\label{fig:coverChange}
\end{figure}

@@

>>>>

The continuously recorded soil moisture data also showed treatment effects on soil moisture, but these effects were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}). We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.  Treatment differences were slightly larger during rainy periods, especially in the spring. We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.

The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest on average. Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature. During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content recorded by the data loggers.

@@

The cover of \textit{H. comata} and \textit{P. spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}). The cover of \textit{P. secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).  In contrast to the grasses, the cover of \textit{A. tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).

Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species. After comparing WAIC scores for the growth models with and without the size by treatment effects, we retained size by treatment interaction effects in the growth models for \textit{A. tripartita} and \textit{P. secunda}, and the survival model for \textit{P. secunda}. For \textit{A. tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls. \textit{A. tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size. Drought led to a strong (but not signficant) decrease in \textit{H. comata} growth, while irrigation had no effect on growth.  Like \textit{A. tripartita}, we saw size by treatment effects on \textit{P.secunda} growth, with the negative effects of drought becoming greater for larger plants. \textit{P. secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although these effects were technically not significant). \textit{P. spicata} growth was relatively unaffected by the drought and irrigation treatments.

Drought decreased the survival of all three grass species (fig \ref{fig:survivalTreat}). And \textit{P. secunda} showed a negative size by drought interaction effect, indicating that the survival of larger plants was more negatively affected by drought than that of the smaller plants, with the smallest plants (plants one standard deviation smaller than the mean) actually seeing a slightly positive effect of drought. \textit{A. tripartita} survival was relatively unaffected by the drought and irrigation treatments.

Recruitment in our irrigation plots was significantly less than in control plots for two grass species \textit{P. secunda} and \textit{P. spicata} (fig \ref{fig:recruitmentTreat}). However, recruitment was also lower in the drought plots than in the the control plots (although not significantly so), indicating that the decrease in the irrigated plots may have not been entirely due to the irrigation itself. The recruitment data for \textit{A. tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment.

\subsection*{Soil moisture effects}

We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the year effects in the random year effects model fit to the observational data (table \ref{table:strongCor}). We included size by soil moisture variables for \textit{A. tripartita} and  \textit{P. secunda} based on the treatment response we observed in the experiment. All three time lags and all three seasons show up in the choosen variables. After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses \ref{fig:climateGrowth}. For \textit{H.comata} the soil moisture of the most recent summer had a significantly positive effect while the soil moisture of the previous summer and the fall before that were also positive but not significant. For \textit{A. tripartita} the summer and fall soil moisture of the previous year had strong negative effects on growth.  There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants.

Soil moisture had significant effects on the survival of all four species \ref{fig:climateSurvival}. As for growth the grasses showed mainly positive effects while \textit{A. tripartita} showed a significant negative effect of soil moisture in the summer of the previous year and a strong negative effect of spring soil moisture of the previous year. \textit{H. comata} showed a significant positive effect of spring lag soil moisture and a strong positive effect of spring soil moisture in the previous year. \textit{P. secunda} showed a significant positive effect of the previous spring's soil moisture and there was a significant interaction between this effect and plant size: as plant size increased this effect became more strongly positive. Finally for \textit{P. spicata} there was a significant positive effect of lag spring soil moisture on survival.

There were only two significant effects of soil moisture on recruitment: lag fall soil moisture had a positive effect on \textit{P. secunda}, and lag summer soil moisture had a negative effect on \textit{P. spicata} recruitment. Soil moisture of the previous year's summer had also had a strong negative effect on \textit{P. spicata} recruitment.

@@

\section*{Tables}

@@

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Mon Dec 19 20:47:29 2016
\begin{table}[ht]
	\centering
	\begin{tabular}{lllrrrr}
		\hline
		vital\_rate & species & climate variable & Int. cor. & p val. & Size cor. & Size p. val. \\ 
		\hline
		growth & ARTR & VWC.su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
		growth & ARTR & VWC.f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
		growth & ARTR & VWC.sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
		growth & HECO & VWC.su.1 & 0.69 & 0.00 &  &  \\ 
		growth & HECO & VWC.su.0 & 0.50 & 0.02 &  &  \\ 
		growth & HECO & VWC.f.lag & 0.37 & 0.10 &  &  \\ 
		growth & POSE & VWC.f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
		growth & POSE & VWC.su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
		growth & POSE & VWC.sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
		growth & PSSP & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
		growth & PSSP & VWC.su.lag & 0.25 & 0.27 &  &  \\ 
		growth & PSSP & VWC.f.0 & -0.22 & 0.34 &  &  \\ 
		recruitment & ARTR & VWC.su.lag & -0.32 & 0.16 &  &  \\ 
		recruitment & ARTR & VWC.su.0 & -0.26 & 0.25 &  &  \\ 
		recruitment & ARTR & VWC.sp.1 & 0.22 & 0.34 &  &  \\ 
		recruitment & HECO & VWC.su.lag & -0.31 & 0.18 &  &  \\ 
		recruitment & HECO & VWC.su.0 & -0.30 & 0.18 &  &  \\ 
		recruitment & HECO & VWC.f.lag & 0.19 & 0.40 &  &  \\ 
		recruitment & POSE & VWC.sp.1 & 0.49 & 0.02 &  &  \\ 
		recruitment & POSE & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
		recruitment & POSE & VWC.f.1 & 0.32 & 0.16 &  &  \\ 
		recruitment & PSSP & VWC.su.lag & -0.52 & 0.02 &  &  \\ 
		recruitment & PSSP & VWC.su.0 & -0.48 & 0.03 &  &  \\ 
		recruitment & PSSP & VWC.sp.0 & 0.30 & 0.19 &  &  \\ 
		survival & ARTR & VWC.su.0 & -0.60 & 0.00 &  &  \\ 
		survival & ARTR & VWC.sp.0 & -0.41 & 0.06 &  &  \\ 
		survival & ARTR & VWC.su.1 & -0.40 & 0.07 &  &  \\ 
		survival & HECO & VWC.sp.0 & 0.44 & 0.04 &  &  \\ 
		survival & HECO & VWC.sp.lag & 0.43 & 0.05 &  &  \\ 
		survival & HECO & VWC.f.1 & 0.33 & 0.14 &  &  \\ 
		survival & POSE & VWC.sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
		survival & POSE & VWC.sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
		survival & POSE & VWC.f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
		survival & PSSP & VWC.sp.0 & 0.36 & 0.11 &  &  \\ 
		survival & PSSP & VWC.sp.lag & 0.34 & 0.13 &  &  \\ 
		survival & PSSP & VWC.su.1 & 0.26 & 0.26 &  &  \\ 
		\hline
	\end{tabular}
	\caption{Correlations between intercept of year effects and soil moisture variables. P-values of correlations are given next to correlations. For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.} 
	\label{table:strongCor}
\end{table}

\section*{Figures}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{start_to_finish_cover_change}
	\caption{Log change in cover in each of the experimental plots from the pre-treatment monitoring in 2011 to the last year of the experiment in 2016. Box plots show the median cover change and the interquartile range. ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}.}
	\label{fig:coverChange}
\end{figure}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{climate_effect_growth}
	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the growth of all four species. Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left. Dark blue parameters show size x climate interaction effects. We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.  ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}. }
	\label{fig:climateGrowth}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{climate_effect_survival}
	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species. Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left. Dark blue parameters show size x climate interaction effects. We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.  ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}. }
	\label{fig:climateSurvival}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{climate_effect_recruitment}
	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species. Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left. Dark blue parameters show size x climate interaction effects. We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.  ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}. }
	\label{fig:climateRecruitment}
\end{figure}
